A Fast and Tight Heuristic for A* in Road Networks

### 2020-09-08 paper/ch_potentials.tex
final changes by Ben

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in large road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They are not based on A* and do not use a heuristic but compute auxiliary data during a preprocessing phase.
% Hierarchical techniques are known to work well on road networks.
They achieve speed and exactness but sacrifice A*'s flexibility.
In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute A* heuristics.
We call our technique CH-Potentials.
With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
\end{abstract}

@@

ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
However, ALT's heuristic is not preprocessing-tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value. % du hast mein "the" wieder rausgenommen, aber ohne Artikel klingt das fuer mich einfach falsch... Lasse mich aber gerne eines besseren belehren.
Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
This results in a good but not a preprocessing-tight heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue operations. % queue pushs stimmt hier nicht...
We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
Avoiding pushing low degree nodes into the queue is the focus of this section.

@@

\subparagraph{Experiments}
The first experiment varies the tightness of the heuristic.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
After $\alpha = 1.1$ the growth slows down.
We observe that the running times for a fixed $\alpha$ vary strongly.
This is an interesting observation, as with uniform source and target sampling, nearly all queries are long-distance.
The query distance is thus not the reason.
After some investigation, we concluded that this is due to non-uniform road graph density.
Some regions have more roads per area than others.
% glaub das mit dem ellispoid lenkt eher ab und braeuchte wenn dann noch eine ganze menge extraerklaerungen...
The number explored A* nodes depends on the density of the search space area.
As the density varies, the running times vary.

@@

In this paper, we introduced CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and CH for finding shortest paths in road networks.
The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity.
The performance of CH-Potentials crucially depends on the availability of good lower bounds in the preprocessing phase.
Our experiments show, that this availability highly depends on the application.
We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the preprocessing-tight heuristic.
% We conclude from this, that to achieve significantly smaller running times with A* based algorithms in road networks, variations of the problem setting are necessary.
Achieving significantly smaller running times could still be possible in variations of the problem setting. % hab das mal ein bisschen mehr auf future work hin getrimmt
This leads to multiple avenues for future research.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in large road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
In road networks, shortest paths can alterantively be quickly computed using hierarchical techniques.
They are not based on A* and do not use a heuristic.
Hierarchical techniques use a slow preprocessing and fast query phase.
The preprocessing phase computes auxiliary data that is used in the query phase.
The auxiliary is independent of origin or destination of the search.
% Hierarchical techniques are known to work well on road networks.
Hierarchical techniques achieve speed and exactness but sacrifice A*'s flexibility.
In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute preprocessing-tight A* heuristics.
We call our technique CH-Potentials.
With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
\end{abstract}

@@

ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
However, ALT's heuristic is not preprocessing-tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
This results in a good but not a preprocessing-tight heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue operations.
We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
Avoiding pushing low degree nodes into the queue is the focus of this section.

@@

\subparagraph{Experiments}
The first experiment varies the tightness of the heuristic.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
After $\alpha = 1.1$ the growth slows down.
We observe that the running times for a fixed $\alpha$ vary strongly.
This is an interesting observation, as with uniform source and target sampling, nearly all queries are long-distance.
The query distance is thus not the reason.
After some investigation, we concluded that this is due to non-uniform road graph density.
Some regions have more roads per area than others.
The number explored A* nodes depends on the density of the search space area.
As the density varies, the running times vary.

@@

In this paper, we introduced CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and CH for finding shortest paths in road networks.
The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity.
The performance of CH-Potentials crucially depends on the availability of good lower bounds in the preprocessing phase.
Our experiments show, that this availability highly depends on the application.
We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the preprocessing-tight heuristic.
Achieving significantly smaller running times could still be possible in variations of the problem setting.
This leads to multiple avenues for future research.
